---
title: "STAT 647 HW#2 - Part 2"
author: "Tiffany Chang"
date: "2023-10-06"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries and custom functions

```{r}
## Required libraries
library(fields)
library(MASS)
library(Matrix)
library(RandomFields)
library(geoR) 
library(geostatsp)

## Custom functions
## Note: Comment out *all* example sims in those R files, so that they don't start running when loading the custom 
## functions below!
source("ConditionalLikelihoodEdited.R")
source("Group3_pairwise.R")
source("Group4_tapered_likelihood.R")
```


## Question 2

a - i) For d = 1, uniformly sample the spatial process over the range [0, 2] and n = 10000.

```{r}
## Define parameters
rho <- .75
sigma <- 1 # sigma=sigma^2=1
alpha <- 1 / rho
nu <- 0.5
gamma <- 2
phi <- gamma^2
n  <- 10000
size <- 2

## Generate data
set.seed(42)
coords <- runif(n,0,size) # Over range [0,size]
D <- rdist(coords) # Distance matrix

#################################################################################################################

## Simulations for Model 1 (exponential cov)

## Method 1: Conditional Likelihood
sims1 <- matrix(NA, nrow=1, ncol=1) # Restricting the number of sims, M=1, since my PC cannot handle large sims
## Note: M is supposed to be 400!

## Obtain the estimates and populate sims1
for(i in c(1:1)){ # M=1
  set.seed(i)
  
  Sigma_grid <- Matern(D, alpha = alpha, nu = nu, phi = phi)
  
  z <- rmvnorm(n = 1, mu = rep(5, n), Sigma = Sigma_grid) # Mean: 5; Sigma: cov(ep_1(s_1),ep_1(s_2)); 
  # n = 1 Y(s) output per simulation rep
  
  cov.pars <- c(1) # Initial parameter value
  
  lower_bound <- 0  # Replace with your desired lower bound
  upper_bound <- 4  # Replace with your desired upper bound

  ## Given that vecchia_likelihood <- function(params, y, mu, gamma, sigma, D, damped = F)
  tryCatch ({ # Use tryCatch to skip any iterations which the likelihood function throws an error. Those will be 
  # recorded as NA in the sims1 matrix.
  out1 = optim(par = cov.pars, # par=cov.pars=params
               fn = vecchia_likelihood,
               
               ## These are the additional arguments within the "vecchia_likelihood" function
               y = z, # y=z=Y(s)
               mu = 5,
               gamma = gamma, # gamma=2
               sigma = sigma, # sigma=1
               D = D,
               damped = F, # damped=F=exponential cov option
               ## End of defining additional arguments
               
               method = "L-BFGS-B", # Use optim with L-BFGS-B method
               lower = lower_bound, 
               upper = upper_bound, 
               control = list(maxit = 10000,trace = 1)) # Example max iterations
  
  sims1[i] = out1$par # Store results
  print(i)
  }, error = function(e) {
    sims1[i] <- c(NA)
  })
}

## Obtain distribution, mean, and sd for each param estimate of the above simulation
par(mfrow=c(3,1))
hist(sims1[,1], main ="Histogram 1-D [0, 2], Exponential, n=10000", xlab ="Rho for Exp")
print(paste("Mean of Rho for Exp Cov: ",mean(sims1[,1], na.rm = TRUE)))
print(paste("Bias of Rho for Exp Cov: ",mean(sims1[,1], na.rm = TRUE) - .75))
print(paste("SD of Rho for Exp Cov: ",sd(sims1[,1], na.rm =TRUE)))

## Method 2: Pairwise Likelihood

## Define additional parameters specific to Pairwise Likelihood
n_reduced <- 10 # Restrict sample size to n=10, since my PC might crash with large n
p <- 3 # Number of beta parameters
coords_reduced <- runif(n_reduced,0,size) # Over range [0,size]
D_reduced <- rdist(coords_reduced) # Distance matrix

sims2 <- matrix(NA, nrow=5, ncol=5)

## Generate data with Matern covariance 
## Y=beta'X+e+epsilon=z=Y(s)
## E(e)=0, cov(e)~Matern(D, alpha=1/rho, nu=0.5, phi=4.0), where rho=.75
## epsilon~N(0,sigma^2), where sigma^2=1

## Obtain the estimates and populate sims1
for(i in c(1:5)){ # M=1
  beta0 = t(c(1,4,3)) # Not sure why they set the beta parameters to these values specifically...
  
  set.seed(i)
  
  Sigma_grid <- Matern(D_reduced, alpha = alpha, nu = nu, phi = phi)
  
  X <- matrix(1, nrow = p, ncol = n_reduced) + matrix(rnorm(p*n_reduced, 0, 1), nrow = p, ncol = n_reduced)
  Y <- t(beta0%*%X + rmvnorm(n = 1, mu = 5, Sigma = Sigma_grid) + rnorm(n_reduced, 0, sigma^2)) # Where rmvnorm(n
  # = 1, mu = 0, Sigma = Sigma_grid) is our "z" data
  
  ## Initial parameter values
  beta_initial = t(rep(1,p))
  sigma2_initial = 1 # Initial value of sigma^2=1
  alpha_initial = 1
  param=c(beta_initial, sigma2_initial, alpha_initial)
  
  tryCatch ({
  out1 <- nlm(likelihood.pairwise, param, n = n_reduced, stepmax=5, print.level=2, gradtol=10^(-10)) # Note: I
  # edited group 3's Pairwise Likelihood function to include the parameter, n, to specify the sample size of the 
  # simulated data.
  
  sims2[i,] <- out1$est
  
  ## Convert alpha to rho
  sims2[i,5] <- 1/sims2[i,5]
  
  print(i)
  }, error = function(e) {
    sims2[i,] <- c(NA, NA, NA, NA, NA)
  })
}

## Obtain distribution, mean, and sd for each param estimate of the above simulation
par(mfrow=c(3,1))
hist(sims2[,1], main ="Histogram 1-D [0, 2], Exponential, n=10000", xlab ="Beta_1 for Exp")
hist(sims2[,2], main ="Histogram 1-D [0, 2], Exponential, n=10000", xlab ="Beta_2 for Exp")
hist(sims2[,3], main ="Histogram 1-D [0, 2], Exponential, n=10000", xlab ="Beta_3 for Exp")
hist(sims2[,4], main ="Histogram 1-D [0, 2], Exponential, n=10000", xlab ="Sigma^2 for Exp")
hist(sims2[,5], main ="Histogram 1-D [0, 2], Exponential, n=10000", xlab ="Rho for Exp")

print(paste("Mean of Beta_1 for Exp Cov: ",mean(sims2[,1], na.rm = TRUE)))
print(paste("Mean of Beta_2 for Exp Cov: ",mean(sims2[,2], na.rm = TRUE)))
print(paste("Mean of Beta_3 for Exp Cov: ",mean(sims2[,3], na.rm = TRUE)))
print(paste("Mean of Sigma^2 for Exp Cov: ",mean(sims2[,4], na.rm = TRUE)))
print(paste("Mean of Rho for Exp Cov: ",mean(sims2[,5], na.rm = TRUE)))

print(paste("Bias of Beta_1 for Exp Cov: ",mean(sims2[,1], na.rm = TRUE) - 1))
print(paste("Bias of Beta_2 for Exp Cov: ",mean(sims2[,2], na.rm = TRUE) - 4))
print(paste("Bias of Beta_3 for Exp Cov: ",mean(sims2[,3], na.rm = TRUE) - 3))
print(paste("Bias of Sigma^2 for Exp Cov: ",mean(sims2[,4], na.rm = TRUE) - 1))
print(paste("Bias of Rho for Exp Cov: ",mean(sims2[,5], na.rm = TRUE) - .75))

print(paste("SD of Beta_1 for Exp Cov: ",sd(sims2[,1], na.rm =TRUE)))
print(paste("SD of Beta_2 for Exp Cov: ",sd(sims2[,2], na.rm =TRUE)))
print(paste("SD of Beta_3 for Exp Cov: ",sd(sims2[,3], na.rm =TRUE)))
print(paste("SD of Sigma^2 for Exp Cov: ",sd(sims2[,4], na.rm =TRUE)))
print(paste("SD of Rho for Exp Cov: ",sd(sims2[,5], na.rm =TRUE)))

## Method 3: Tapered Likelihood
## To be continued...
