---
title: "STAT 647 HW#3"
author: "Tiffany Chang"
date: "2023-11-05"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

```{r}
## a) Simulating temporal SAR models

## Make the proximity/adjacency matrix (below we use the one in Example 4.1.1)
set.seed(101)
r1 = c(0,1/3,0,1/3,1/3,0)
r2 = c(1/3,0,1/3,0,1/3,0)
r3 = c(0,1/3,0,0,1/3,1/3)
r4 = c(1/2,0,0,0,1/2,0)
r5 = c(1/5,1/5,1/5,1/5,0,1/5)
r6 = c(0,0,1/2,0,1/2,0)

## Create the Weighted Matrix
Wtilde = matrix(c(r1,r2,r3,r4,r5,r6),ncol=6,byrow=T)s

## Create a function that can simulate a temporal SAR
## Here sigma acts as standard deviation, not variance
temporalSAR = function(rho1,rho2,rho3,sigma,W,TT,burnin){
  # get number of columns of the weight matrix
  n = ncol(W)
  # generate n random values from a N(0, \sigma^2) distribution
  Y = rnorm(n = n, mean = 0, sd = sigma)
  # Create 2 of the same columns of X that were generated in the previous line
  Y = cbind(Y,Y)
  # Create an identity matrix of size n
  In = diag(rep(1,n)) 
  # Calculate (I_n - rho_1 W)^{-1} (see page 87 in notes)
  Arho1 = solve(In - rho1*W)
  # Calculate (rho_2 I_n + rho_3 W) (see page 87 in notes)
  Brho23 = (rho2*In+rho3*W) 
  ArhBrh = Arho1%*%Brho23
  
  for(i in c(3:(TT + burnin))){
    # create a noise parameter sampled from N(0, sigma^2)
    epsilon = rnorm(n = n, mean = 0, sd = sigma)
    # Calculate Y_t (see page 87 eq.4.12)
    temp =  ArhBrh%*%Y[,(i-1)]+Arho1%*%epsilon 
    # attatch new Y_t value to the Y matrix
    Y = cbind(Y,temp)  
    
  }
  
  Yout = Y[,-c(1:burnin)]
  
  return(Yout)
  
}

#############################################################
# Stability is an issue, I think we require that the eigenvalues of
# Arho1%*%Brho23 all lie within the unit circle

# This means you need to choose the parameters wisely.
# If after burn the series explodes just look at the numbers
# then you know the parameters cause the process to unstable.

#############################################################

## Simulation #1: rho1=.3, rho2=.5, rho3=.2, sigma=1
sim1 = temporalSAR(rho1 = 0.3,rho2 = 0.5,rho3 = 0.2,sigma = 1,W = Wtilde,TT = 100,burnin = 200)
ncol(sim1)
nrow(sim1)

## Plot time series at each region (6 total)
par(mfrow=c(2,3))

plot.ts(sim1[1,])
plot.ts(sim1[2,])
plot.ts(sim1[3,])
plot.ts(sim1[4,])
plot.ts(sim1[5,])
plot.ts(sim1[6,])

## Simulation #2: rho1=.3, rho2=.3, rho3=.6, sigma=1
sim2 = temporalSAR(rho1=0.3,rho2=0.3,rho3=0.6,sigma=1,W=Wtilde,TT=100,burnin=100)
ncol(sim2)
nrow(sim2)

par(mfrow=c(2,3))

plot.ts(sim2[1,])
plot.ts(sim2[2,])
plot.ts(sim2[3,])
plot.ts(sim2[4,])
plot.ts(sim2[5,])
plot.ts(sim2[6,])
```
In the first SAR model, the time series look very similar for each region, which means that its neighborhoods do not have a large impact on Y. It does appear that there is cyclic movements, where Y peaks at around Time=50 and dips at around Time=80 and Time=15, which means that there is large temporal impact on Y. However, the cycle does not repeat at regular intervals (non-periodic) [i.e. the peaks and dips happen at unequally spaced time points]. This result makes sense since rho1 (=.3) and rho3 (=.2) are the correlation coefficients associated with the impact of the region's neighbors in the present (Time=t) and past (Time=t-1), respectively, and they both have relatively weaker correlation compared to that of time alone (rho2=.5). (*Please verify that I did not misinterpret anything here.)

Source #1: https://www.jstor.org/stable/2276710?seq=1
Source #2: https://support.minitab.com/en-us/minitab/21/help-and-how-to/graphs/time-series-plot/interpret-the-results/key-results/
Source #3: https://www.investopedia.com/terms/s/serial-correlation.asp#:~:text=Serial%20correlation%20occurs%20in%20a,variable%20affects%20its%20future%20level.

It appears that for the second SAR model the parameters caused the process to become unstable (non-stationary over time) since the Y values of the time series increased dramatically nearing Time=100 (*Not sure if that is what I should discuss in my observation or if it's just something wrong with the code.)
